The galactic chemical evolution of carbon: Constraining stellar nucleosynthesis

### 2023-03-21 carbon_paper/tex/main.tex
more plot fixes

A fundamental question of Galactic Chemical Evolution (GCE) is understanding the physical processes which make each element. Stellar evolution models can be used to predict how much of each element stars with a variety of physical properties produce. However, these models are rife with uncertainties, limited by our understanding of critical physical processes such as nuclear reaction rates, convection, opacity, and mass loss (CITE, e.g. stellar physics reviews).

@@

Galactic chemical evolution are fundamentally intertwined with our
understanding of stellar nucleosynthesis. In this paper, we are concerned with
three major nucleosynthetic pathways: AGB stars, \CC, and Supernovae type Ia.
While only AGB stars and \CC are producers of carbon, SNeIa Fe can be used to
compare with the AGB delayed carbon component.

\VICE allows us to explore the nucleosynthesis of various models, as it
includes AGB tables for C11, et.c

We discuss carbon nucleosythesis in more detail in the next two sections. We
assume that the only dominant sources of carbon are from AGB stars and CCSNe/massive star winds. ***JUSTIFY***. Below  is a brief overview of our  nucleosythetic assumptions of other elements.

Let $y_{\rm X}^{\rm proc}(M, Z)$ be the fractional yield of element X by a process proc for stars of a mass $M$ (in solar masses) and at metallicity $Z$ defined as: 
\begin{equation}
    y_{\rm X} = \frac{M_{\rm X,\ ej} - M_{\rm X,\ birth}}{M}   
\end{equation}

Where $M_{\rm X,\ ej}$ and $M_{\rm X,\ birth}$ are the mass of element X ejected and initially present in the star. 
We define the IMF-integrated yield of X with: 
\begin{equation}
Y_{\rm X}^{\rm proc}(Z) = \int_{M_{\rm min}}^{M_{\rm max}} y_{\rm X}^{\rm proc}(M, Z)\ \Phi(M)  \ dM
\end{equation}
Where $\Phi(M) = \frac{dN}{dM}/ \int_{M_{\rm min}}^{M_{\rm max}} \frac{dN}{dM}\ dM$ is the normalized IMF, $M_{\rm min}$ and $M_{\rm max}$ are the minimum and maximum mass of stars, which we take to be $0.08 M_{\sun}$ and $100 M_{\sun}$ respectively; and $\frac{dN}{dM}$ is the initial mass function, the frequency of stars born between $M$ and $M+dM$.

There are two $\alpha$-elements we care about: Mg and O. Stellar observations of magnesium are more reliable but oxygen abundances are easier to measure in HII regions. We assume that both Mg and O are produced entirely in CCSNe with constant, metallicity-independent yields. Thus $\text{[Mg/H]} = \text{[O/H]} = [\alpha/\text{H}]$ in our models

\begin{itemize}
    \item observational validation
\end{itemize}
For carbon, the dominant producer is from massive stars which eject carbon fused in their cores during CCSNe.

Taken from J21 and J22, we make the following yield assumptions:
\begin{itemize}
    \item $Y_\text{O}^\text{CC} = 0.015$
    \item $Y_\text{O}^\text{Ia} = 0$
    \item $Y_\text{Fe}^\text{Ia} = 0.00214?$
    \item $Y_\text{Fe}^\text{CC} = 0.0012$
    \item $Y_\text{N}^\text{CC} = 0.00036$
    \item $y_\text{N}^\text{AGB}(M, Z) = 9\times 10^{-4} M \left(\frac{Z}{Z_\odot}\right)$
\end{itemize}
Also following J21/J22, we take the SNeIa Delay Time distribution to be a $t^{-1.1}$ law based on
the observations by \citet{maoz+12}.

We leave consistent throughout the paper. While different CCSNe models predict different absolute values for $y_\text{O}^\text{CC}$, they also predict that the yield is metallicity-independent within the uncertainties of the model.

@@

An AGB star is a low mass ($\lesssim 8 M_{\sun}$) star during the final phases of evolution. The star has already completed Hydrogen core burning, Hydrogen shell burning during the red giant branch phase, and Helium core burning. So, the structure consists of an inert core of carbon and oxygen surrounded by a Helium burning shell and a Hydrogen burning shell. The double shell burning leads to thermal instabilities which drive off the outer layers of the star in a series of violent pulsations, ejecting the envelope and leaving behind a C/O white dwarf.

In an AGB star, TDU and HBB are two major processes involved in carbon production and destruction. TDU occurs during thermal pulsations, where the convective layer of the star passes into the inner layers with He-burning products which mixes in C and O into the envelope. In contrast, HBB may happen if the base of the convective envelope reaches about 50MK, initiating the CNO cycle, converting most C12 into N14.

@@

    \includegraphics[scale=1]{y_agb_vs_z-1.pdf}
    \includegraphics[scale=1]{y_agb_vs_t-1.pdf}

@@

@@

We summarise many of the predictions of stellar models in Fig. \ref{fig:y_cc}. Pink squares
represent \citep{NKT13}, etc. etc.
\citep{LC18}
\citep{sukhbold+16}
\citep{WW95}

On the right, we plot the expected CCSNe abundance ratio of each model as a function of metallicity, defining
\begin{equation}
    {\rm [C/O]^{CC}} = \log_{10}\left( \frac{y_{\rm C}^{\rm CC}}{y_{\rm O}^{\rm CC}}\right) - \log_{10} \left( \frac{Z_{C, \sun }}{Z_{O, \sun }} \right)
\end{equation}
Notice how even these ratios 
\begin{figure*}
    \includegraphics[scale=0.9]{y_c_cc-2.pdf}
    \includegraphics[scale=0.9]{y_co_cc.pdf}
    \caption{The fist plot shows IMF-weighted net fractional CCSNe yield of carbon, $y_C^{cc}$ against metalicity, [M/H], for different studies. The right plot instead shows [C/O]$^\text{CC}$, the logarithm of the abundance ratio of carbon to oxygen if CCSNe were the only carbon source, also as a function of metallicity for the same set of studies. The black line represents the carbon yield of the fiducial model, $y_C^{CC} = 0.004? (Z/Z_{\sun})^{0.3}$, and becomes dashed over regions where we have not tested our model. }
    \label{fig:y_cc}
\end{figure*}

@@

\section{The Multizone Model}

@@

\subsection{Metallicity dependence and the AGB Fraction}
The next question is how does changing the CCSNe carbon yield affect the model?
We now consider variations of our fiducial model:
\begin{itemize}
    \item $f_{\rm agb} = 0.2$
    \item AGB model: \cristallo
    \item $Z_{\rm C}^{\rm cc} = 0.004 (Z/Z_{\sun })^{0.4}$ ($\beta = 4$)
\end{itemize}

@@

\begin{figure*}
\centering
\includegraphics[]{summary-2.pdf}
\caption{Present-day gas phase tracks of the fiducial model in \caah~space. M101 is valid comparison to MW ...(van Dokkum et al. 2014)}
\label{fig:gas_phase}
\end{figure*}
\subsection{Outflows}
A common degeneracy in chemical evolution models is the yield-outflow degeneracy. An increase in stellar yields has a near-identical effect as a decrease in outflows. However, reducing the outflows reduces the evolution timescale of the system.

@@

@@

>>>>

Where does each chemical element come from?

Stellar evolution models can be used to predict how much of each element stars with a variety of physical properties produce. However, these models are rife with uncertainties, limited by our understanding of critical physical processes such as nuclear reaction rates, convection, opacity, and mass loss (CITE, e.g. stellar physics reviews).

@@

Modern theoretical nucleosynthesis provides our starting point for our models.  We focus on three primary nucleosynthetic pathways: AGB stars, CCSne, and SNeIa. Each process has unique timescales and element production which we can trace through the tools of GCE. While carbon is only known to be produced in AGB stars and CCSNe, comparing carbon to iron production (from SNeIa) adds a valuable constraint to our models and insight into the delay-time production of carbon.

After a group of stars are formed, CCSNe from massive stars are the first enrichment, providing light elements such as C, O, and Mg and heavier elements such as Fe and beyond. Importaintly, $\alpha$ elements like O and Mg are only formed in CCSNe (in significant quantities). Next, low mass stars begin to reach the end of their lives. The dying breaths of AGB stars are known to be importaint sources of C, N, and a smattering of heavier elements.  Finally, the white dwarfs, including formation and time to collide/over-eat, explode, releasing much of Fe and nearby elements.

We focus on O and Mg as $\alpha$ elements only formed in CCSNe, and use Fe to
trace delayed enrichment from SNeIa. C and N are both formed in AGB and CCSNe.

We use IMF-Integrated yields calculated with \VICE~throughout this paper (see
section \ref{sec:vice}, \citet{james+21, james+22}). For an element X and star
with mass $M$, the stellar yield $y$ is defined as the net production of X relative 
to $M$, or
\begin{equation}
    y_{\rm X} = \frac{M_{\rm X,\ ejected} - Z_{0, X} M_{\rm ejected}}{M}   
\end{equation}
where we let $M_{\rm ejected}$ and $M_{\rm X, ejected}$ be the total ejected
stellar mass (through winds and supernovae ejecta) and the ejected mass of X,
and $Z_{0, X}$ was the birth mass fraction of element X.

More useful for GCE are Initial Mass Function (IMF) yields, where we sum up the
contributions of all stars which produce an element.
We define the IMF-integrated yield of X with: 
\begin{equation}
Y_{\rm X}^{\rm proc}(Z) = \int_{M_{\rm min}}^{M_{\rm max}} y_{\rm X}^{\rm proc}(M, Z)\ \Phi(M)  \ dM
\end{equation}
Where $\Phi(M) = \frac{dN}{dM}/ \int_{M_{\rm min}}^{M_{\rm max}} \frac{dN}{dM}\ dM$ is the normalized IMF, $M_{\rm min}$ and $M_{\rm max}$ are the minimum and maximum mass of stars, which we take to be $0.08 M_{\sun}$ and $100 M_{\sun}$ respectively; and $\frac{dN}{dM}$ is the initial mass function, the frequency of stars born between $M$ and $M+dM$.

There are two $\alpha$-elements we care about: Mg and O. Stellar observations
of magnesium are more reliable but oxygen abundances are easier to measure in
HII regions. We assume that both Mg and O are produced entirely in CCSNe with
constant, metallicity-independent yields. Thus $\text{[Mg/H]} = \text{[O/H]} =
[\alpha/\text{H}]$ in our models. (Observational validation?)

For carbon, the dominant producer is from massive stars which eject carbon fused in their cores during CCSNe.

As we focus on the yields of carbon, we keep yields for other elements fixed in
most models. Taken from \citet{james+21, james+22}, we let
\begin{itemize}
    \item $Y_\text{O}^\text{CC} = 0.015$
    \item $Y_\text{O}^\text{Ia} = 0$
    \item $Y_\text{Fe}^\text{Ia} = 0.00214?$
    \item $Y_\text{Fe}^\text{CC} = 0.0012$
    \item $Y_\text{N}^\text{CC} = 0.00036$
    \item $y_\text{N}^\text{AGB}(M, Z) = 9\times 10^{-4} M \left(\frac{Z}{Z_\odot}\right)$
\end{itemize}
Also following \citet{james+21, james+22}, we take the SNeIa Delay Time distribution to be a
$t^{-1.1}$ law suggested by the observations of \citet{maoz+12}.

@@

    \includegraphics[scale=1]{y_agb_vs_z.pdf}
    \includegraphics[scale=1]{y_agb_vs_t.pdf}

An AGB star is a low mass ($\lesssim 8 M_{\sun}$) star during the final phases of evolution. The star has already completed hydrogen core burning, Hydrogen shell burning during the red giant branch phase, and helium core burning. So, the structure consists of an inert core of carbon and oxygen surrounded by a helium burning shell and a hydrogen burning shell. The double shell burning leads to thermal instabilities which drive off the outer layers of the star in a series of violent pulsations, ejecting the envelope and leaving behind a C/O white dwarf.

In an AGB star, two competing processes determine the outcome of carbon production.
Third Dredge Up (TDU) is needed to eject carbon from the star, enriching the surface with helium-burning-processed material from the core. But Hot Bottom Burning (HBB) at the base of the convection layer changes carbon into nitrogen. 
TDU occurs during thermal pulsations, where the convective layer of the star passes into the inner layers with He-burning products which mixes in C and O into the envelope. In contrast, HBB may happen if the base of the convective envelope reaches about 50MK, initiating the CNO cycle, converting most C12 into N14.

@@

@@

\begin{figure*}
    \includegraphics[scale=0.9]{y_c_cc-2.pdf}
    \includegraphics[scale=0.9]{y_co_cc.pdf}
    \caption{The fist plot shows IMF-weighted net fractional CCSNe yield of
        carbon, $y_C^{cc}$ against metalicity, [M/H], for different studies.
        The right plot instead shows [C/O]$^\text{CC}$, the logarithm of the
        abundance ratio of carbon to oxygen if CCSNe were the only carbon
        source, also as a function of metallicity for the same set of studies.
        The black line represents the carbon yield of the fiducial model,
        $y_C^{CC} = 0.004? (Z/Z_{\sun})^{0.3}$, and becomes dashed over regions
        where we have not tested our model.\citep{NKT13}, \citep{LC18},
        \citep{sukhbold+16}, \citep{WW95}
 }
    \label{fig:y_cc}
\end{figure*}

@@

We summarise many of the predictions of stellar models in Fig. \ref{fig:y_cc}. 
\begin{equation}
    {\rm [C/O]^{CC}} = \log_{10}\left( \frac{y_{\rm C}^{\rm CC}}{y_{\rm O}^{\rm CC}}\right) - \log_{10} \left( \frac{Z_{C, \sun }}{Z_{O, \sun }} \right)
\end{equation}
Notice how even these ratios

@@

    \section{The Multizone Model}\label{sec:vice}

@@

\subsection{Metallicity dependence and the AGB Fraction}

The next question is how does changing the CCSNe carbon yield affect the model?
We now consider variations of our fiducial model:
\begin{itemize}
    \item $f_{\rm agb} = 0.2$
    \item AGB model: \cristallo
    \item $Z_{\rm C}^{\rm cc} = 0.004 (Z/Z_{\sun })^{0.4}$ ($\beta = 4$)
\end{itemize}

@@

\subsection{Outflows}
A common degeneracy in chemical evolution models is the yield-outflow degeneracy. An increase in stellar yields has a near-identical effect as a decrease in outflows. However, reducing the outflows reduces the evolution timescale of the system.

@@

\subsection{Understanding Scatter}

Beyond simply comparing median trends, we can compare the distribution 
shape of our distribution to observations.

@@

\begin{figure*}
\centering
\includegraphics[]{summary-2.pdf}
\caption{Present-day gas phase tracks of the fiducial model in \caah~space. M101 is valid comparison to MW ...(van Dokkum et al. 2014)}
\label{fig:gas_phase}
\end{figure*}
